---
title: "Compared to what?"
description: |
  "At the heart of quantitative reasoning is a single question: Compared to what? - Edward Tufte [@tufte1990]"
author:
  - name: "Graphics principles"
    url: https://graphicsprinciples.github.io/
date: "`r Sys.Date()`"
categories:
  - comparisons
  - line plot
  - bar chart
  - Rcode
  - example
draft: FALSE
bibliography: lit.bib
output:
  radix::radix_article:
    self_contained: FALSE
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(gridExtra)

# Set global theme
theme_set(theme_light())

# create a few themes for plotting. 
th <- theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y = element_blank()
)
thg <- theme(
  panel.grid.major = element_line(color = "grey"),
  panel.grid.minor = element_line(color = "grey")
)
th2 <- theme(axis.title.x = element_blank(),
             axis.title.y = element_blank())
th3 <- theme(
  axis.title.x = element_blank(),
  axis.text.x = element_blank(),
  axis.title.y = element_blank()
)
th4 <- theme(
  axis.title.y = element_blank(),
  axis.text.y = element_blank(),
  axis.title.x = element_blank()
)

# Set seed for reproducibility 
set.seed(1972)
```


Effective visualizations communicate complex statistical and quantitative information facilitating insight, understanding, and decision making. But what is an effective graph? It is one that enables visual comparisons of data within the reviewers eye span, and adapts to the audience in a manner that best allows the key message to be understood. 

During the design it is helpful to keep in mind the purpose – the *"Compared to what?"* [@tufte1990].

To facilitate effective comparisons there are a number of design options which can be considered such as layering, alignment, proximity and ordering. Below is a selection of tips to consider during the design phase.


# Visual hierarchy adds context

Utilising visual anchors such as grid and reference lines, annotations and other visual anchors can draw attention to the comparison of interest. To illustrate the point, consider the graph below which is a stripped down version of a scatter plot. 

```{r, fig.cap= "This is an example of a barebones scatter plot."}
# generate data
set.seed(1971)
N <- 100
plotdata <- tibble(x = rnorm(N, 0, 1),
                   y = rnorm(N, x, 1))

# plot x vs y 
plotdata %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point(size = 3) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
    )
```


If you are quantitively trained you may compare the relationship between the x- and y-variables assessing if there is a (linear) relationship by eye. You may also look at the distribution of observations to check for possible "outliers". 

We can start to layer on visual anchors to meet specific goals (the purpose of the graph). For example, adding grid lines and axes help identify the x- and y-values for specific observations. 

```{r, fig.cap="Add grid lines"}
plotdata %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point(size = 2, shape = 19, alpha = 0.35) 

```

Adding a reference line helps distinguish between observations above and below the line. Adding a reference line can also facilitate assessments comparing the distance between the reference line and each observation. 

```{r, fig.cap = "Adding a reference line."}
plotdata %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_hline(yintercept = 0.5, linetype = 1, color = "red", size = 2, alpha = 0.3) +
  geom_point(size = 2, shape = 19, alpha = 0.35) 

```

We can also add a regression line to help assess the association between the x- and y- variables. In this case we add a spline. We can also add an indicator of uncertainty (in this case the standard error). This further helps determine the fit of the regression model with respect to each observation (i.e. the residuals). 

```{r, fig.cap = "Adding a regression"}

plotdata %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_hline(yintercept = 0.5, linetype = 1, color = "red", size = 2, alpha = 0.3) +
  geom_point(size = 2, shape = 19, alpha = 0.35) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3)) 
```



```{r, preview=TRUE, fig.cap= "We can add additional visual anchors to the plot to help guide the comparisons."}
plotdata %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_hline(yintercept = 0.5, linetype = 1, color = "red", size = 2, alpha = 0.3) +
  geom_point(size = 2, shape = 19, alpha = 0.35) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3)) + 
  annotate(x = 1.75, y = 0.3, geom = 'text', label = "Reference line of interest") +
  annotate(x = 1.75, y = -3.4, geom = 'text', label = "This observation is important") +
  annotate("segment", x = 0.8, y = -3.4, xend = -0.1, yend = -3.4,
             arrow = arrow(angle = 30, length = unit(1, "mm"), type = "closed")) +
  xlab("Covariate [unit]") +
  ylab("Outcome [unit]") 

```

Finally we can add additional annotations such as a description of the reference line and highlighting a specific observation. We can also add labels for the x and y- variables adding in additional context. 

The final example illustrates that by layering visual anchors we can guide the focus of the reviewer to specific comparisons of interest such as the association between the outcome and a covairate of interest, the strength of the association (model fit) and also highlight specific observations of interest). 




# Horizontal vs. vertical alignment

Another consideration to faciltiate comparisons is alignment. For example, a vertical comparison is often easier to make judgments on than a horizontal comparison. To illustrate, consider three point estimates and associate uncertainty intervals. 

```{r}
alignData <- tibble(
  x = factor(c(1, 2, 3)),
  x_lab = c("Long label one", "Long label two", "Long label three"),
  y = c(2, 1, 3),
  low = c(1, 0, 2),
  hi = c(3, 2, 4)
  )

a <- alignData %>% ggplot(aes(x = x, y = y, ymin = low, ymax = hi)) + 
  geom_point(size = 7) + 
  geom_linerange(size = 1) + 
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  ggtitle("Horizontal")


b <- alignData %>% 
  ggplot(aes(x = x, y = y, ymin = low, ymax = hi)) + 
  geom_point(size = 7) + 
  geom_linerange(size = 1) + 
  coord_flip() + 
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
    ) +
  ggtitle("Vertical")


grid.arrange(a, b, nrow = 1)
```

When appropriate, a vertical or “downward” comparison between groups helps with assessments especially when comparing across groups. This comes with the caveat that if the comparison is sequential or temporal (i.e. a comparison of longiatidnal effects), doing so would not make sense. 

Vertical alignment of comparisons can also help with adding in contextual informaiton such as labels. 
```{r}

a <- alignData %>% ggplot(aes(x = x_lab, y = y, ymin = low, ymax = hi)) + 
  geom_point(size = 7) + 
  geom_linerange(size = 1) + 
  theme_light(base_size = 16) + 
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    ) +
  ggtitle("Horizontal")


b <- alignData %>% 
  ggplot(aes(x = x_lab, y = y, ymin = low, ymax = hi)) + 
  geom_point(size = 7) + 
  geom_linerange(size = 1) + 
  theme_light(base_size = 16) + 
  coord_flip() + 
  theme(
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
    ) +
  ggtitle("Vertical")


grid.arrange(a, b, nrow = 1)

```



# Overlaying

When comparing few factors (e.g. 2-3), consider overlaying in the same plot. 

```{r}
n <- 100
set.seed(5)
overlay <- data.frame(v=c(1,2), mu=c(1,1.4))
overlay <- overlay[rep(c(1,2),n),]
overlay$s <- rnorm(n=2*n,mean=overlay$mu, sd=1)
ggplot(overlay, aes(x=s)) + facet_wrap(~v, ncol=1) + geom_line(stat="density",size=1) + th + thg
ggplot(overlay, aes(x=s, colour=factor(v))) + geom_line(stat="density", size=1) + 
  scale_color_brewer(palette="Dark2", name="") + th +thg
```

**Changes**: The two distributions are overlaid on the same graph and color added for further differentiation.  

## Distance

Keep things that are to be compared close together. 

```{r}
distance <- data.frame(expand.grid(study=c("Study 1","Study 2"), trt=c("A","B")))
distance$y <- c(1.2, 1.1, 1.8, 1.9)
ggplot(distance, aes(x=trt, y=y, ymin=0, ymax=y)) + 
  facet_wrap(~study, ncol=2) + geom_linerange(size=10) + 
  th2 + thg
ggplot(distance, aes(x=study, y=y, ymin=0, ymax=y)) + 
  facet_wrap(~trt, ncol=2) + geom_linerange(size=10) + 
  th2 + thg
```

**Changes**: Study moved to x-axis and treatment group to the panel to allow comparison within treatment across study.  Note: if the comparison of interest is A vs. B within study then the left plot would be more appropriate; keep in mind the “compared to what?”

## Ordering

Ordering by values presented is helpful for comparing many categories. 

```{r}
ordering <- data.frame(v=LETTERS[1:7], y=runif(7))
ggplot(ordering, aes(x=y, y=v)) + geom_point(size=5) + th3 + thg
ordering <- arrange(ordering, y)
ggplot(ordering, aes(x=y, y=factor(v, levels=ordering$v))) + geom_point(size=5) + th3 + thg
```

**Changes**: Factors ordered by observed value, largest to smallest.

# Difference – do the math

Show the difference where applicable (e.g. plot the mean difference rather than two means) to avoid the reviewer doing so.  

```{r} 
difference <- data.frame(grp=c("PBO", "TRT1", "TRT2"), y=c(1,2,2.2), low=c(0,1,1.1), hi=c(2,3,3.1))
ggplot(difference, aes(x=grp, y=y, ymin=low, ymax=hi)) + geom_pointrange(size=1) + th3 +  thg + coord_flip()
difference2 <- data.frame(grp=c("TRT1 - PBO","TRT2 - PBO"), y=c(1,1.2), low=c(-0.3,-0.1), hi=c(2.3,2.5))
ggplot(difference2, aes(x=grp, y=y, ymin=low, ymax=hi)) + geom_hline(yintercept=0, linetype=2) + 
  geom_pointrange(size=1) + scale_y_continuous(breaks=c(0,1,2)) + coord_flip() + th3 +thg
```

**Changes**: The difference to placebo is computed and shown directly for each treatment group.

Another example with longitudinal data

```{r}
# another difference example with longitudinal data. 
set.seed(2)
rm(difference)
difference <- data.frame(expand.grid(grp=c("A","B"), day=c(1,7,14,21,28))) %>% 
  arrange(grp, day)
difference <- difference %>% mutate(true_mean=ifelse(grp=="A", 0.5, -0.5), 
         y=rnorm(dim(difference)[1], true_mean, 0.55)) 
ggplot(difference, aes(x=day, y=y, colour=grp)) + 
  geom_point(size=2.5) + geom_line(size=1, alpha=0.5) + 
  scale_x_continuous(breaks=unique(difference$day)) + 
  scale_colour_brewer(palette="Dark2", name="") + labs(x="Time", y="Outcome") + 
  theme(axis.text.x=element_blank()) +thg

difference2 <- spread(select(difference, day, grp, y), grp, y) %>%
  mutate(difference=A-B)
ggplot(difference2, aes(x=day, y=difference)) + 
  geom_point(size=2.5) + geom_line(size=1, alpha=0.5) + 
  geom_hline(yintercept=0, linetype=3, colour="black") + 
  scale_x_continuous(breaks=unique(difference$day)) + 
  labs(x="Time", y="A - B") + theme(axis.text.x=element_blank()) + thg
```
