---
title: "Compared to what?"
description: |
  "Part 2: alignment and easing visual inspection
author:
  - name: "Graphics principles"
    url: https://graphicsprinciples.github.io/
date: "`r Sys.Date()`"
categories:
  - comparisons
  - alignment
  - Rcode
  - examples
draft: FALSE
bibliography: lit.bib
output:
  radix::radix_article:
    self_contained: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


During the design of data visualisation it is helpful to keep in mind the purpose – the *"Compared to what?"* [@tufte1990]. To facilitate effective comparisons there are a number of design options in the toolbox which can be used including layering, alignment, proximity and ordering. Following on from the [previous post](https://graphicsprinciples.github.io/gallery/posts/2019/2019-09-18-compared-to-what/), we now focus on how the alignment of a comparison is an important consideration. 


```{r}
library(tidyverse)
library(gridExtra)

# Set global ggplot theme
theme_set(theme_light())

# Set seed for reproducibility 
set.seed(1972)

# generate data
alignData <- tibble(
  x = factor(c(1, 2, 3)),
  x_lab = c("Long label one", "Long label two", "Long label three"),
  y = c(2, 1, 3),
  low = c(1, 0, 2),
  hi = c(3, 2, 4))
```


# Horizontal vs. vertical alignment

Alignment is another consideration for facilitating comparisons. Where appropriate, using a vertical comparison can help make judgments easier than the same comparisons horizontally aligned. To illustrate, consider three point estimates and associated uncertainty intervals. 

```{r}
a <- alignData %>%
  ggplot(aes(
    x = x,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  ggtitle("Horizontal")


b <- alignData %>% 
  ggplot(aes(
    x = x,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  coord_flip() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  ggtitle("Vertical")

# plot side by side
grid.arrange(a, b, nrow = 1)
```

Back to the when appropriate caveat. If the comparison is made on data that is sequential (i.e. a comparison of longitudinal effects), doing so may confuse the reviewer; time is often encoded better by flowing left to right. 

Vertical alignment of comparisons can also help with adding in contextual information such as labels. 

```{r}

# add in axis labels
a <-
  alignData %>% ggplot(aes(
    x = x_lab,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  theme_light(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  ggtitle("Horizontal")


b <- alignData %>%
  ggplot(aes(
    x = x_lab,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  theme_light(base_size = 16) +
  coord_flip() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  ggtitle("Vertical")


grid.arrange(a, b, nrow = 1)

```


On the left-hand side the labels blend in together. Typically this requires shortening the label length through abbreviation or by other means such as changing the angle the text direction. By switching the alignment of the comparison (i.e. flipping the coordinates), the longer labels are easier to display without modification that may impact readability.  

## Corrections or questions {.appendix}

If you see mistakes, want to suggest changes or even ask us a question, please [create an issue](https://github.com/GraphicsPrinciples/gallery/issues) on the source repository.
