---
title: "Compared to what?"
description: |
  "Part 2: alignment
author:
  - name: "Graphics principles"
    url: https://graphicsprinciples.github.io/
date: "`r Sys.Date()`"
categories:
  - comparisons
  - alignment
  - Rcode
  - examples
draft: FALSE
bibliography: lit.bib
output:
  radix::radix_article:
    self_contained: FALSE
    toc: true
    toc_depth: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


During the design of data visualisation it is helpful to keep in mind the purpose – the *"Compared to what?"* [@tufte1990].

To facilitate effective comparisons there are a number of design options in the toolbox which can be used including layering, alignment, proximity and ordering. In this post we focus on how the alignment of a comparison is an important consideration. 


```{r}
library(tidyverse)
library(gridExtra)

# Set global ggplot theme
theme_set(theme_light())

# Set seed for reproducibility 
set.seed(1972)

# generate data
alignData <- tibble(
  x = factor(c(1, 2, 3)),
  x_lab = c("Long label one", "Long label two", "Long label three"),
  y = c(2, 1, 3),
  low = c(1, 0, 2),
  hi = c(3, 2, 4))
```


# Horizontal vs. vertical alignment

Alignment is another consideration for facilitating comparisons. To be more specific, a vertical comparison is often easier to make judgments on than a horizontal comparison. To illustrate, consider three point estimates and associated uncertainty intervals. 

```{r}
a <- alignData %>%
  ggplot(aes(
    x = x,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  ggtitle("Horizontal")


b <- alignData %>% 
  ggplot(aes(
    x = x,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  coord_flip() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  ggtitle("Vertical")

# plot side by side
grid.arrange(a, b, nrow = 1)
```

When appropriate, a vertical or “downward” comparison between groups helps with assessments especially when comparing across groups. 

This comes with the caveat that if the comparison is sequential or temporal (i.e. a comparison of longitudinal effects), doing so may not make sense to the reviewer; time is often encoded better by flowing left to right. 

Vertical alignment of comparisons can also help with adding in contextual information such as labels. 

```{r}

# add in axis labels
a <-
  alignData %>% ggplot(aes(
    x = x_lab,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  theme_light(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  ggtitle("Horizontal")


b <- alignData %>%
  ggplot(aes(
    x = x_lab,
    y = y,
    ymin = low,
    ymax = hi
  )) +
  geom_point(size = 7) +
  geom_linerange(size = 1) +
  theme_light(base_size = 16) +
  coord_flip() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  ggtitle("Vertical")


grid.arrange(a, b, nrow = 1)

```


By switching the alignment of the comparison, the longer labels are able to be displayed. 

## Corrections or questions {.appendix}

If you see mistakes, want to suggest changes or even ask us a question, please [create an issue](https://github.com/GraphicsPrinciples/gallery/issues) on the source repository.
